---
title: "Global FoodOmics - GI Tube dataset"
author: "Kiana A West"
date: "`r Sys.Date()`"
params:
  metadata: "../data/ST001794_GI_tube_dataset_metadata.csv"
  foods: "../data/GI_tube_food_consumption.csv"
  f_counts: "../data/GI_tube_food_counts_L5.csv"
  save_dir: "../output/GI_Tube"
output: 
  html_document:
    toc: true
    theme: cerulean
---

<style type="text/css">
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
</style>

***  

This data was collected from the small intestine of a single person over ~8 hours. Foods consumed during the study are listed below.

```{r, echo=FALSE}
options(stringsAsFactors = FALSE)
```

Make output directory

```{r}
if (!dir.exists(params$save_dir)) { system(paste("mkdir", params$save_dir)) }
```

Load packages

```{r}
pacman::p_load("magrittr", "ggpubr", "tidyverse")
```

Read in (meta)data

```{r, message=FALSE}
metadata <- read_csv(params$metadata)
f_counts <- read_csv(params$f_counts)
foods <- read_csv(params$foods)
```

Foods consumed during the study

```{r, echo=FALSE}
knitr::kable(foods)
foods$`Sampling Time` <- str_replace_all(foods$`Sampling Time`, c(":00$" = "", ":" = ".")) %>%
  as.numeric() %>%
  round(digits = 2)
```

Transform counts to proportions

```{r}
f_counts_norm <- f_counts %>%
  dplyr::select(-filename) %>%
  apply(1, function(x) {ifelse(x != 0, (x/sum(x))*100, 0)}) %>%
  t() %>%
  data.frame() %>%
  mutate(filename = f_counts$filename)
```

Foods to plot

```{r}
gfop_foods <- c("apple", "coffee", "pear", "peach", "pumpkin", "cinnamon", "olive", "wheat", "yogurt_cow")
```

Merge data and metadata

```{r}
to_plot <- f_counts_norm %>%
  select(one_of(gfop_foods, "filename")) %>%
  left_join(metadata) %>%
  mutate(time_point = as.numeric(gsub("Time point ", "", time_point))) %>%
  left_join(foods, by = c("time_point" = "Sample Number")) %>%
  pivot_longer(all_of(gfop_foods), names_to = "food", values_to = "proportion")
```

Plot

```{r, fig.align='center'}
(p <- ggline(to_plot, x = "Sampling Time", y = "proportion", group = "food", color = "food",
             ylab = "Proportion of spectral matches (%)", numeric.x.axis = TRUE) +
   theme(axis.text = element_text(size = 7)) +
   # manually add times foods were eaten on the plot
   annotate(geom = "text", x = 9.35, y = 7.6, label = "apple juice, coffee", size = 2.5, angle = 90) +
   annotate(geom = "text", x = 10.45, y = 7.6, label = "puree", size = 2.5, angle = 90) +
   annotate(geom = "text", x = 10.55, y = 7.6, label = "puree", size = 2.5, angle = 90) +
   annotate(geom = "text", x = 12.10, y = 7.6, label = "olive oil, bread, coffee", size = 2.5, angle = 90) +
   annotate(geom = "text", x = 14.15, y = 7.6, label = "cereal, coffee", size = 2.5, angle = 90) +
   annotate(geom = "text", x = 16.35, y = 7.6, label = "kefir", size = 2.5, angle = 90))
```

```{r}
ggsave(filename = file.path(params$save_dir, "GI_tube_time_series.pdf"), plot = p)
```

## Session Info

```{r, echo=FALSE}
sessionInfo()
```

