---
title: "Global FoodOmics - U19 - ADNI"
subtitle: "Correlations between foods (L5) and cognition"
author: "Kiana A West"
date: "`r Sys.Date()`"
params:
  f_counts_file: "../data/ADNI_food_counts_L5.csv"
  meta_file: "../data/ADNI_metadata_KAW.csv"
  imp_var: "../data/ADNI_GO_Variables_Of_Importance.csv"
  imp_foods: "../output/ADNI/ADNI_imp_foods.csv"
  save_dir: "../output/ADNI"
output: 
  html_document:
    toc: true
    theme: cerulean
---

<style type="text/css">
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
</style>

***  

Make output directory

```{r}
if (!dir.exists(params$save_dir)) { system(paste("mkdir", params$save_dir)) }
```

Load packages

```{r}
pacman::p_load("plotly", "tidyverse", "magrittr", "ggpubr")
```

Read in files

```{r read_inputs, message=FALSE}
meta <- read_csv(params$meta_file) %>%
  filter(RID != "RID999999", # remove non-clinical samples
         !duplicated(RID)) # use only one sample per patient
f_counts <- read_csv(params$f_counts_file) %>%
  filter(filename %in% meta$SampleID) # keep only samples with metadata
imp_var <- read_csv(params$imp_var) # variables of interest
imp_foods <- read_csv(params$imp_foods) # important foods identified in overall diet analysis
```

Transform food counts to proportions

```{r, message=FALSE, warning=FALSE}
transform_f_counts <- function(f) {
  fc <- read_csv(f)
  filenames <- fc$filename
  fc <- fc %>%
    dplyr::select(-filename) %>%
    apply(., 1, function(x) {ifelse(x != 0, x/sum(x), 0)}) %>%
    t() %>%
    data.frame() %>%
    set_rownames(make.names(filenames))
  return(fc)
}

f_counts <- transform_f_counts(params$f_counts_file) %>%
  select(-mouse, -worm) %>% # remove zoo foods
  select(one_of("filename", imp_foods$food)) # use important foods identified in overall diet analysis
f_counts <- f_counts[row.names(f_counts) %in% meta$SampleID,]
```

Order metadata like f_counts

```{r}
meta <- left_join(data.frame(SampleID = row.names(f_counts)), meta) %>%
  set_rownames(row.names(f_counts))
```

# Are certain foods related to cognition?

Model data using Partial Least Squares (PLS) regression

```{r, fig.align='center', fig.height=8, fig.width=8}
m <- mixOmics::pls(f_counts,
                   select(meta, RAVLT_immediate, RAVLT_learning, RAVLT_forgetting, RAVLT_perc_forgetting),
                   mode = "canonical")
mixOmics::cim(m, transpose = T, row.cex = 0.5, col.cex = 0.6)
```

```{r, echo=FALSE}
# save plot
mixOmics::cim(m, transpose = T, save = "pdf", name.save = file.path(params$save_dir, "PLS_foods_L5_cognition"), 
              row.cex = 0.5, col.cex = 0.6)
```

# Are certain foods related to variables of interest?

Most variables of interest are missing from the metadata we received.

```{r}
imp_var$Variable
```

```{r, fig.align='center', fig.height=8, fig.width=8}
df <- meta %>%
  select(DX_bl, AGE, PTGENDER, PTEDUCAT, APOE4, TAU) %>%
  mutate(DX_bl = case_when(DX_bl == "CN" ~ 0,
                           DX_bl == "SMC" ~ 1,
                           DX_bl == "EMCI" ~ 2,
                           DX_bl == "LMCI" ~ 3,
                           DX_bl == "AD" ~ 4),
         PTGENDER = case_when(PTGENDER == "Male" ~ 1,
                              PTGENDER == "Female" ~ 0))
m <- mixOmics::pls(f_counts, 
                   df,
                   mode = "canonical")
mixOmics::cim(m, transpose = T, row.cex = 0.5, col.cex = 0.6)
```

```{r, echo=FALSE}
# save plot
mixOmics::cim(m, transpose = T, save = "pdf", name.save = file.path(params$save_dir, "PLS_foods_L5_imp_var"), 
              row.cex = 0.5, col.cex = 0.6)
```

# Session Info

```{r}
sessionInfo()
```

