---
title: "Global FoodOmics - U19"
subtitle: "BEAM cross-over diet: changes in diet profiles (L4 & L5) over time"
author: "Kiana A West"
date: "`r Sys.Date()`"
params:
  ont_file: "../../foodomics_metadata_08APR21.tsv"
  f_counts_dir: "../data"
  f_counts_name: "BEAM_.*_L4|BEAM_.*_L5"
  meta_file: "../data/BEAM_metadata_KAW.csv"
  save_dir: "../output/BEAM"
output: 
  html_document:
    toc: true
    theme: cerulean
---

<style type="text/css">
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
</style>

***  

Participants completed one diet (Keto or American Heart Association - AHA), did a wash-out, then switched to the other diet. We wanted to see if the diet readout reflects this.

***   
   
Make output directory

```{r}
if (!dir.exists(params$save_dir)) { system(paste("mkdir", params$save_dir)) }
```

Load packages

```{r}
pacman::p_load("magrittr", "ggpubr", "mixOmics", "santaR", "tidyverse")
```

Read in files

```{r read_inputs, message=FALSE}
ont <- read_delim(params$ont_file, delim = "\t") %>%
  filter(!duplicated(sample_name)) # two food samples have the same name "mackerel_gonad_13"
meta <- read_csv(params$meta_file)
# get files matching string provided
f_counts_files <- list.files(params$f_counts_dir, pattern = params$f_counts_name)
for (f in f_counts_files) {
  assign(gsub("\\.csv", "", f), read.csv(file.path(params$f_counts_dir, f), row.names = 1))
}
```

Format data

```{r}
ont <- set_rownames(as(ont, "matrix"), ont$sample_name)
meta <- set_rownames(as.data.frame(meta), meta$filename)
```

Transform food counts to proportions

```{r}
transform_f_counts <- function(fc) {
  filenames <- row.names(fc)
  fc <- fc %>%
    apply(., 1, function(x) {ifelse(x != 0, x/sum(x), 0)}) %>%
    t() %>%
    data.frame() %>%
    mutate(filename = filenames) %>%
    left_join(meta, by = "filename")
  return(fc)
}

for (f in grep("^BEAM_.*food_counts", ls(), value = TRUE)) {
  transf <- transform_f_counts(get(f))
  assign(paste0("prop_", f), transf)
}
```

Transform food counts using centered log-ratio (CLR)

```{r}
for (f in grep("^BEAM_.*food_counts", ls(), value = TRUE)) {
  transf <- logratio.transfo(get(f), logratio = "CLR", offset = 1) %>%
    as("matrix") %>%
    data.frame() %>%
    rownames_to_column(var = "filename") %>%
    left_join(meta, by = "filename")
  assign(paste0("clr_", f), transf)
}
```

# Can we see longitudinal changes in diet profiles?

Time series analysis on individual foods using spline-fitting method implemented in [santaR](https://cran.r-project.org/web/packages/santaR/santaR.pdf)

```{r, spline_fit, message=FALSE, warning=FALSE}
# get levels used from environment
levs <- grep("food_counts_L", ls(), value = TRUE) %>%
  gsub(".*food_counts_", "", .) %>%
  unique()
for (transf in c("prop", "clr")) { # analyze both transformations
  for (lev in levs) { # and all levels provided
    for (biosp in c("fecal", "serum")) { # and both biospecimen types
      df <- get(paste0(transf, "_BEAM_", biosp, "_food_counts_", lev))
      # keep patients with at least 4 time points
      keep <- names(which(table(df$SubjectID) > 3))
      df <- df %>%
        filter(First.Diet %in% c("1", "2"),
               SubjectID %in% keep)
      foods <- colnames(get(paste0("BEAM_", biosp, "_food_counts_", lev)))
      sp <- santaR_auto_fit(inputData = df[,foods], ind = df[["SubjectID"]], 
                            time = as.numeric(df[["Study_TP"]]), group = df[["First.Diet"]], 
                            df = 4)
      assign(paste("sp", transf, lev, biosp, sep = "_"), sp)
    }
  }
}
```

Plot mean spline for each group for each food

```{r}
plot_spline <- function(sp, savename) {
  foods <- names(sp)
  # get BH-corrected pvalues
  pvals <- santaR_auto_summary(sp)$pval.all %>%
    mutate(food = row.names(.))
  for (f in 1:length(foods)) {
    food <- foods[f]
    pval <- round(pvals$dist_BH[pvals$food == food], digits = 4)
    # plot group mean splines with confidence intervals
    p <- santaR_plot(sp[[food]], 
                     showIndPoint=FALSE,
                     showIndCurve=FALSE,
                     xlab = "Time point", 
                     ylab = ifelse(grepl("clr", savename), "CLR(food count)", "Food abundance (%)"), 
                     title = paste(food, paste0("(FDR=", pval, ")")),
                     colorVect = c("orange", "dodgerblue")) +
      scale_x_continuous(labels = c("1" = "1: Pre-diet 1", "2" = "2: Diet 1", "3" = "3: Post-diet 1/Pre-diet 2", 
                                    "4" = "4: Diet 2", "5" = "5: Post-diet 2")) +
      rotate_x_text(angle = 45) +
      theme(axis.text = element_text(size = 5),
            panel.grid.major = element_blank(), 
            panel.grid.minor = element_blank(), 
            title = element_text(size = 7)) 
    ggsave(file.path(params$save_dir, paste0("BEAM_time_series_plot_", savename, "_", food, ".pdf")), 
           p, width = 4, height = 3)
    # if p<0.05, show plot in notebook
    if (pvals$dist[pvals$food == food]<0.05) { print(p) }
  }
}
```

Plot will be displayed if `P<0.05`, even if not significant after adjusting for multiple hypothesis tests. `Group 1` started with Keto diet, `Group 2` started with AHA diet.

```{r, fig.align='center'}
for (sp in grep("sp_", ls(), value = TRUE)) {
  obj <- get(sp)
  suppressMessages(plot_spline(obj, gsub("sp_", "", sp)))
}
```

# Session Info

```{r}
sessionInfo()
```