---
title: "Global FoodOmics - IBD"
subtitle: "Exploratory analysis of all datasets"
author: "Kiana A West"
date: "`r Sys.Date()`"
params:
  ibd_200: "../data/IBD_200.RDS"
  ibd_1: "../data/IBD_1.RDS"
  ibd_biospy: "../data/IBD_biopsy.RDS"
  ibd_seed: "../data/IBD_seed.RDS"
  ibd_seed2: "../data/IBD_seed2.RDS"
  save_dir: "../output"
output: 
  html_document:
    toc: true
    theme: cerulean
---

<style type="text/css">
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
</style>

***  
Here we analyzed untargeted metabolomics data from 5 IBD cohorts to get an idea of variation in dietary patterns. Food counts (pseudo diet readouts) were generated from clinical samples according to the following (workflow)[https://ccms-ucsd.github.io/GNPSDocumentation/tutorials/rdd/].
***   

Make output directory

```{r}
if (!dir.exists(params$save_dir)) { system(paste("mkdir", params$save_dir)) }
```

Load packages

```{r}
pacman::p_load("phyloseq", "magrittr", "ggpubr", "plotly", "mixOmics", "tidyverse", "UpSetR")
```

Read in data

```{r read_inputs}
ibd_200 <- readRDS(params$ibd_200)
ibd_1 <- readRDS(params$ibd_1)
ibd_seed <- readRDS(params$ibd_seed)
ibd_biopsy <- readRDS(params$ibd_biospy)
ibd_seed2 <- readRDS(params$ibd_seed2)
```

Combine into one phyloseq object

```{r}
ps_all <- merge_phyloseq(ibd_200$ps, ibd_1$ps, ibd_seed$ps, ibd_biopsy$ps, ibd_seed2$ps)
```

Summary table

```{r}
sumtab <- data.frame(ps_all@sam_data, stringsAsFactors = FALSE) %>%
  group_by(Cohort, Study.sub.group) %>%
  dplyr::summarise(samples = n(),
                   individuals = n_distinct(host_subject_id)) %>%
  mutate(n = paste0(samples, " (", individuals, ")")) %>%
  dplyr::select(-samples, -individuals) %>%
  pivot_wider(names_from = "Study.sub.group", values_from = "n")
write.csv(sumtab, file.path(params$save_dir, "dataset_summary_table.csv"))
```

```{r, echo=FALSE}
knitr::kable(sumtab)
```

Note: `NA` samples are cases where metadata was missing.

# PCA of all datasets

```{r}
f_counts_mat <- as(ps_all@otu_table, "matrix")
clr_mat <- logratio.transfo(f_counts_mat, offset = 1, logratio = "CLR")
m <- pca(clr_mat, 
         ncomp = 3, center = T, scale = T)
df <- data.frame(m$variates$X, filename = m$names$sample, total_sum = rowSums(f_counts_mat), 
                 unique_foods = apply(f_counts_mat, 1, function(x) {sum(x > 0)}))
df <- left_join(data.frame(sample_data(ps_all), stringsAsFactors = FALSE), df, by = "filename")
df <- subset(df, !is.na(df$Study.sub.group))
```

Plot PCA scores (interactive plot is saved as html)

```{r}
for (f in c("Cohort", "Study.sub.group")) {
  p <- plot_ly(df, x = ~PC1, y = ~PC2, type = "scatter", mode = "markers", color = ~get(f),
               colors = c("slategrey", "aquamarine2", "khaki", "darkgoldenrod", "brown"),
               marker = list(size = 10, width = 2, line = list(color = "black", width = 2), opacity = 1),
               text = ~filename) %>%
    layout(title = f,
           xaxis = list(title = paste0("PC1", " [", round(m$prop_expl_var$X[1]*100, digits = 1), "%]"), size = 18),
           yaxis = list(title = paste0("PC2", " [", round(m$prop_expl_var$X[2]*100, digits = 1), "%]"), size = 18))
  htmlwidgets::saveWidget(p, file = file.path(params$save_dir, paste0("PCA_food_counts_colorBY_", f, ".html")))
}
```

Plot PCA scores (static plot for notebook)

```{r, fig.align='center'}
for (f in c("Cohort", "Study.sub.group")) {
  p <- ggscatter(df, x = "PC1", y = "PC2", color = f, alpha = 0.8,
                 palette = c("slategrey", "aquamarine2", "khaki", "darkgoldenrod", "brown"),
                 xlab = paste0("PC1", " [", round(m$prop_expl_var$X[1]*100, digits = 1), "%]"),
                 ylab = paste0("PC2", " [", round(m$prop_expl_var$X[2]*100, digits = 1), "%]")) +
    theme(axis.text = element_text(size = 7)) +
    geom_hline(yintercept = 0, lty = 2, size = 0.2) +
    geom_vline(xintercept = 0, lty = 2, size = 0.2)
  print(p)
}
```

# Most common foods found in fecal vs biopsy samples

```{r}
to_plot <- data.frame(ps_all@sam_data, stringsAsFactors = FALSE) %>%
  dplyr::select(Cohort, ends_with("L5"))

# most prevalent foods in each cohort
prev <- to_plot %>%
  group_by(Cohort) %>%
  summarise_all(function(x) {sum(x > 0)/length(x)}) %>%
  dplyr::select(-mouse_L5)

top_foods <- names(sort(colSums(prev[,-1], na.rm = T), decreasing = T))[1:30]
```
 
Plot
 
```{r, fig.align='center'}
to_plot <- to_plot %>%
  dplyr::select(Cohort, one_of(top_foods)) %>%
  pivot_longer(-Cohort, names_to = "Food", values_to = "RelativeAbundance") %>%
  mutate(Food = str_replace_all(Food, c("_L5" = "", "_" = " ")))
(p <- ggviolin(to_plot, x = "Cohort", y = "RelativeAbundance", add = "jitter", 
              add.params = list(color = "Cohort", alpha = 0.3), 
              palette = c("slategrey", "aquamarine2", "khaki", "darkgoldenrod", "brown")) +
  labs(y = "Relative Abundance") +
  facet_wrap(. ~ Food, scales = "free_y") +
  rotate_x_text(angle = 45) +
  theme(axis.text = element_text(size = 6), legend.position = "none", strip.text = element_text(size = 8)))
ggsave(filename = file.path(params$save_dir, "relative_abundances_most_prevalent_foods.pdf"), width = 8)
```

# Session Info

```{r, echo=FALSE}
sessionInfo()
```